ELAVL1 KO analysis in irradiated MCF7 and MB231 cell lines
========================================================

```{r setup, echo=FALSE, include=FALSE, cache=FALSE}
library(knitr) 
opts_chunk$set(cache.path='cache/', fig.path='img/', cache=F, tidy=T, fig.keep='high', echo=F, dpi=300, out.width=700)
options(replace.assign=TRUE, width=120)
rname<-"results//"
```

```{r loadData, echo=FALSE, warning=FALSE}
X<-as.matrix(read.table("data//data_all.txt", sep="\t", header=T))
class(X)<-"numeric"
fNames<-readLines("data//featureNames.txt") # miRNA names
fType<-readLines("data//featureType.txt") # Targets and Housekeeping
# pheno Data frame of conditions
P = data.frame(Cell=factor(ifelse(grepl("231", tolower(colnames(X))), "MB231", "MCF7")),
               Rad=factor(ifelse(grepl("5gy", tolower(colnames(X))), "Rpos", "Rneg")),
               KD=factor(ifelse(grepl("sihur", tolower(colnames(X))), "Sih", "Scr")))

```

The data were normalized by *Endogenous control*, then using quantile normalization. The Ct values were converted to expression counts using 2^(-Ct) formula.

```{r preprocessData}
# Normalize by deltaCt 
normalizeCt = function(X, ref.genes) {
  Xn = X
  for (c in 1:ncol(Xn)) {
    Xn[,c] = X[,c] - mean(X[ref.genes, c], na.rm=T)
  }
  Xn
}

library(limma)
X[is.na(X)] = max(X, na.rm=T) # Replace undetected by maximum
Xn = normalizeQuantiles(normalizeCt(X, grep("Endogenous Control", fType))) # Normalize by Ct and quantile
rownames(Xn)<-fNames; colnames(Xn)<-colnames(X)
Xne = 2 ^ (- Xn[!(fType %in% "Endogenous Control"), ]) # Convert to copy number counts
```

We visualize the top 20 most highly exressed miRNAs, and their relative expression level, simply for an overview.

```{r visualization, fig.width=4, fig.height=4, error=FALSE}
suppressMessages(library(gplots))
rank.cor.dist = function(X) {
  ranks = t(apply(X, 2, function(x){rank(x, ties.method="average")}))
  as.dist((1-cor(ranks)) / 2)
}
#par(oma=c(5,0,0,5), mar=c(10, 4.1, 4.1, 5), cex.main=0.25)
par(cex.main=0.25)
heatmap.2(-Xne[order(-rowSums(Xne))[2:22],], col=redgreen(70), Colv=F, distfun=rank.cor.dist, 
          scale="row", density.info="none", trace="none", cexRow=0.5, cexCol=0.5, 
          main="Top 20 highly expressed miRNAs, Red/reen - high/low")
```

Principal component analysis for each cell type shows we have both siRNA and Rad effects.

```{r pcaMCF7, fig.width=3, fig.height=3}
par(cex=0.5)
pcaVisual<-function(cellType){
  X.pca = Xne[,P$Cell==cellType]
  pca = prcomp(X.pca)$rotation
  x = pca[,1]
  y = pca[,2]
  xadj<-0.1*(max(x)-min(x))
  yadj<-0.1*(max(y)-min(y))
  plot(x,y,xlab="PC1", ylab="PC2", main=cellType,
       xlim=c(min(x) - xadj, max(x) + xadj), ylim=c(min(y) - yadj, max(y) + yadj),
       pch=ifelse(P$KD=="Sih", 1, 2),
       col=ifelse(P$Rad=="Rpos", "red", "blue")) 
  text(x,y+0.05,labels=rownames(pca))
  legend("bottomright", c("Sih/Rpos", "Scr/Rpos", "Sih/Rneg", "Scr/Rneg"), 
       col=c("red","red","blue","blue"),
       pch=c(1,2,1,2))
}
pcaVisual("MCF7")

pcaVisual("MB231")
```

For each miRNA we define a ratio of siRNA effect in non-irradiated vs. irradiated cells, referred hereafter as the ratio.

```{r differences, echo=FALSE}
Rad_KO_Interaction<-function(cellType, trimOutliers=FALSE){
  a<-(Xne[,P$Cell==cellType & P$KD=="Sih" & P$Rad=="Rpos"] /
      Xne[,P$Cell==cellType & P$KD=="Scr" & P$Rad=="Rpos"]) /
      (Xne[,P$Cell==cellType & P$KD=="Sih" & P$Rad=="Rneg"] /
      Xne[,P$Cell==cellType & P$KD=="Scr" & P$Rad=="Rneg"]) 
  for (i in 1:length(a)){
    if (a[i]<1) {a[i]<- -(1/a[i])} # Convert rations to +/- scale
  }
  a.out<-a
  fname<-cellType
  if (trimOutliers) {
    a<-a[!a==max(a)]
    a<-a[!a==min(a)]
    fname<-paste(fname, "trimmed", sep="_")
  }
  cut.min<-summary(a)[4]-1*sd(a, na.rm=T) # Low 1SD cutoff
  cut.max<-summary(a)[4]+1*sd(a, na.rm=T) # High 1SD cutoff

  par(cex=0.5)
  hist(a, n=length(unique(a)), freq=F, main=paste("Density distribution of the ratio for", cellType), xlab="Normalized expression counts")
  curve(dnorm(x, mean=mean(a), sd=sd(a)), add=T )
  
  cat("Min threshold:", cut.min, "how many miRNAs:", sum(a<cut.min), 
      "Max threshold:", cut.max, "how many miRNAs:", sum(a>cut.max), "\n")
  
  mtx<- -Xne[a.out<cut.min | a.out>cut.max, P$Cell==cellType]
  heatmap.2(mtx, Colv=F, col=redgreen(70), scale="row", density.info="none", trace="none", cexRow=0.5, cexCol=0.5)
  write.table(a.out[a.out<cut.min | a.out>cut.max], paste(rname, fname, ".txt", sep=""), sep="\t", row.names=T, col.names=F)
  cat("Raw Ct values\n")
  print(cbind(rownames(mtx), X[fNames %in% rownames(mtx), P$Cell==cellType]))
  cat("\nNormalized Ct values\n
      Note that the mean of endogenous control probes was subtracted from the Ct values, hence some values are negative. Look for RELATIVE comparisons, not absolute Ct values.\n")
  print(formatC(Xne[rownames(Xne) %in% rownames(mtx), P$Cell==cellType], format="f"))
  return(rownames(mtx))
}
```

MCF7
----------------

First, we look at the distribution of the ratio. Then, we fit a normal distribution to it, and define min/max cutoffs as 1 standard deviation around its mean. We check how many miRNAs are above the maximum or below the minimum thresholds and visualize their normalized expression counts as a heatmap. We also check the raw Ct values for these miRNAs.

```{r noTrimmingMCF7, fig.width=4, fig.height=4, results='markup', warning=FALSE, error=FALSE}
c1.1<-Rad_KO_Interaction("MCF7", trimOutliers=F)
```

We observe two outliers on both sides of the distribution. They expand the fitting of the normal distribution, and make min/max thresholds overinflated. 

To alleviate this bias, we remove those outliers, check the cutoffs and the number of miRNAs above the cutoffs, visualize them as a heatmap and look at their raw Ct values. 

```{r withTrimmingMCF7, fig.width=4, fig.height=4, results='markup', warning=FALSE, error=FALSE}
c1.2<-Rad_KO_Interaction("MCF7", trimOutliers=T)
```

MB231
-------------

We repeat the same steps for MB231 cell line, first using the data as is,

```{r noTrimmingMB231, fig.width=4, fig.height=4, results='markup', warning=FALSE, error=FALSE}
c2.1<-Rad_KO_Interaction("MB231", trimOutliers=F)
```

then, trimming outliers.

```{r withTrimmingMB231, fig.width=4, fig.height=4, results='markup', warning=FALSE, error=FALSE}
c2.2<-Rad_KO_Interaction("MB231", trimOutliers=T)
```


```{r vennDiagPrepare}
source("venn4.R")

combineMatrix<-function(i, j){
  m.uniq<-unique(c(i, j))
  mtx<-matrix(data=0, nrow=length(m.uniq), ncol=2)
  rownames(mtx)<-m.uniq
  mtx[i, 1]<-1
  mtx[j, 2]<-1
  mtx
}
```

Overlap between cell lines
--------------------------
First, check overlap between the most significant miRNAs (without removing outliers).
```{r vennTop}
mtx<-combineMatrix(c1.1, c1.2)
colnames(mtx)<-c("MCF7", "MB231")
vennDiagram(mtx)
r.mtx<-rownames(mtx)[rowSums(mtx) == 2]
par(oma=c(4,0,0,3))
heatmap.2(-Xne[r.mtx,], col=redgreen(70), distfun=rank.cor.dist, Colv=F,
          scale="row", density.info="none", trace="none", cexRow=1, cexCol=1, 
          main="Overlapping top miRNAs\nRed/reen - high/low")
cat("\nNormalized RELATIVE Ct value changes\n")
print(Xn[r.mtx,])
writeLines(r.mtx, paste(rname, "MCF7-MB231_top", ".txt", sep=""))
```

Then, check overlap between the larger lists of significant miRNAs (with removing outliers).
```{r vennTop+}
mtx<-combineMatrix(c2.1, c2.2)
colnames(mtx)<-c("MCF7", "MB231")
vennDiagram(mtx)
r.mtx<-rownames(mtx)[rowSums(mtx) == 2]
par(oma=c(4,0,0,3))
heatmap.2(-Xne[r.mtx,], col=redgreen(70), distfun=rank.cor.dist, Colv=F,
          scale="row", density.info="none", trace="none", cexRow=1, cexCol=1, 
          main="Overlapping top+ miRNAs\nRed/reen - high/low")
cat("\nNormalized RELATIVE Ct value changes\n")
print(Xn[r.mtx,])
writeLines(r.mtx, paste(rname, "MCF7-MB231_top+", ".txt", sep=""))
```


